Predicting Political Support in the United States

### 2024-03-16 output/paper.qmd
data update w/ cross ref, paper.pdf is outdated, need to fix latex $$ issue

  The topic of concern is the 2020 United States presidential election held on November 3rd, 2020, where we saw Democratic Party leader Joe Biden win against incumbent president Donald Trump (Gambino 2020). Using the 2022 Cooperative Election Study (CES) survey data, we aim to predict who a voter supported in the 2020 election based on their cultural background, political stances, and other factors. We will be using a representative sample of 60000 Americans to discern patterns in these voters' responses  and how their personal backgrounds may have influenced their voting decision. We will be examining the impact of various factors such as: gender, race, educational background, religon, and personal stances on key political issues such as gun control and student loans. Our main focus is to compare the correlations in the 2022 CES survey data to the 2020 CES survey data, and to see if our estimand, the effect of a voter's cultural background and political stances on their voting decision, has influenced their voting decision in the 2020 election. Using these understandings, we hope to highlight the potential implications these correlations may have for this upcoming 2024 election.

@@

* `ggplot2` [@citeggplot] 
* `dplyr` [@citedplyr] 
* `readr` [@citereadr] 
* `tibble` [@citetibble]

@@

@@

-   `gun_own`, renamed to `gun_ownership`,

@@

@@

    It was important to get the voter's `age`, in order to understand the generational differences and how their experiences and personal interests may influence their voting decision. In our given age buckets, it was discovered that 65% of 18-29 year olds supported the Democrats, whereas only 40% of 65+ year olds supported the Democrats, however, 18-29 year olds only made up 12% of the sample, whereas 65+ year olds made up 30% of the sample.

@@

    Incorporating the voter's `religion`, was a needed factor since it is important to understand that religious beliefs and differences can have a strong influence one's personal beliefs and therefore impact their personal stances on topics such as gun ownership. To put a contrasting example out there, 30% of Protestant/Christian respondents support the Democrats as opposed to a whopping 72% by Atheists. This could be attributed to certain parties aligning more to certain religious beliefs.

    A voter's `education`, was important because there may be certain understandings they arrive at during the completion of their education. Perhaps they viewed the world in one way at first, but after becoming more 'educated' in a topic, they feel that they have a better understanding of the world and either become more confident in their initial understanding or change their stance entirely. This goes hand in hand with the voter's stance on student loans, `student_loan`, as it is important to understand how a voter's personal experiences with student loans may influence their voting decision. There was a time of serious consideration during Biden's tenure where he was considering forgiving student loans, and this intent alone may have swayed some support in his favour.

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Votes for Candidate given Age Group and Gender"
#| fig-subcap: ["Male and Female", "Other (Non-binary and Other)"]
#| label: fig-agevsreligion
#| layout-ncol: 2

@@

@@

# Appendix {-}

@@

\newpage
# References

>>>>

The topic of concern is the 2020 United States presidential election held on November 3rd, 2020, where we saw Democratic Party leader Joe Biden win against incumbent president Donald Trump [cite2020Election]. Using the 2022 Cooperative Election Study (CES) survey data, we aim to predict who a voter supported in the 2020 election based on their cultural background, political stances, and other factors. We will be using a representative sample of 60000 Americans to discern patterns in these voters' responses and how their personal backgrounds may have influenced their voting decision. We will be examining the impact of various factors such as: gender, race, educational background, religon, and personal stances on key political issues such as gun control and student loans. Our main focus is to compare the correlations in the 2022 CES survey data to the 2020 CES survey data, and to see if our estimand, the effect of a voter's cultural background and political stances on their voting decision, has influenced their voting decision in the 2020 election. Using these understandings, we hope to highlight the potential implications these correlations may have for this upcoming 2024 election.

@@

-   `ggplot2` [@citeggplot]
-   `dplyr` [@citedplyr]
-   `readr` [@citereadr]
-   `tibble` [@citetibble]

@@

Personal Background:

@@

    A small sample of these variables can be seen in the @tbl-cesdata1, which shows rows of respondent data, including their `voted_for`, `age_group`, `gender`, `education`, `race`, and `religion`.

Personal Stances:

-   `gun_own`, renamed to `household_gun_ownership`,

@@

    A small sample of these variables can be seen in the @tbl-cesdata2, which shows rows of respondent data, including their `voted_for`, `household_gun_ownership`, and `student_loan_status`.

@@

    It was important to get the voter's `age`, in order to understand the generational differences and how their experiences and personal interests may influence their voting decision. In our given age buckets, it was discovered that 65% of 18-29 year olds supported the Democrats, whereas only 40% of 65+ year olds supported the Democrats, however, 18-29 year olds only made up 12% of the sample, whereas 65+ year olds made up 30% of the sample. In @fig-agevsgender, we can see how respondents' support for the Democrats and Republicans can differ based on their `age` and `gender`. For example, in @fig-agevsgender-1 'Males' aged 65+ were quite a bit more supportive towards the Republicans as opposed to 'Males' aged 18-29, who showed a noticibly larger gap in support towards the Democrats. In @fig-agevsgender-2, despite the lower respondent count, we can see how 'Other' gendered respondents had significantly more voters in the 18-29, 30-44 age groups, while also showing a larger gap in support towards the Democrats.

@@

    Incorporating the voter's `religion`, was a needed factor since it is important to understand that religious beliefs and differences can have a strong influence one's personal beliefs and therefore impact their personal stances on topics such as gun ownership. To put a contrasting example out there, 30% of Protestant/Christian respondents support the Democrats as opposed to 72% by Atheists. This could be attributed to certain parties aligning more to certain religious beliefs. To see a bigger insight into this, take a look at @fig-racevsreligion1 and @fig-racevsreligion2 (figures separated according to respondent count), and notice examples like, 'White Protestants' appear significantly more supportive towards the Republicans as opposed to 'White Catholics' and 'Black Protestants' who appear to be significantly more supportive towards the Democrats.

    A voter's `education`, was important because there may be certain understandings they arrive at during the completion of their education. Perhaps they viewed the world in one way at first, but after becoming more 'educated' in a topic, they feel that they have a better understanding of the world and either become more confident in their initial understanding or change their stance entirely. @fig-gunownvseducation shows how `education` goes hand in hand with the voter's stance on gun ownership, `household_gun_ownership`, while @fig-loanvseducation shows how `education` goes hand in hand with the voter's stance on student loans, `student_loan`. It is important to understand how a voter's personal experiences with gun ownership and student loans may influence their voting decision. To highlight a key example, there was a time of serious consideration during Biden's tenure where he was considering forgiving student loans [@citeballotpedia], and this intent alone may have swayed some support in his favour.

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "Votes for Candidate given Age Group and Gender"
#| fig-subcap: ["Male and Female", "Other (Non-binary and Other)"]
#| label: fig-agevsgender
#| layout-ncol: 2

@@

\newpage

@@

# Appendix {.unnumbered}

@@

\newpage

# References
